\mytitle

### 2012-02-22 papers/SAS2012/implicitization.tex
Ça rentre presque dans les 15 pages

\usepackage{amsmath}
\usepackage{mathcomp,amsfonts,amssymb,mathabx}
\usepackage[mathscr]{euscript}
\usepackage{algorithmic,algorithm}
\usepackage{paralist}
\usepackage{graphicx}
%\usepackage{color}

@@

Most SMT solvers, including Z3,
%\footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
Yices,
%\footnote{\url{http://yices.csl.sri.com/}}
and all those supporting the full SMTLIB2 standard \cite{BarST-SMTLIB},
offer an \emph{incremental} interface: the client program specifies the formula as an initially empty conjunction, to which additional constraints are added, and calls a ``check'' function answering whether it is satisfiable; 
it may then backtrack some of the constraints and add other ones without restarting from scratch.

@@

Classical interval analysis on this program yields $\lstinline|x_old| \in [-100000,100000]$.
We have transformed the first program into the second, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first \avirer{(as in Fig.~\ref{fig:multigraph})}, even though not all feasible paths are needed to compute the invariant.

Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit while simulating its analysis. This means we work on an implicitly represented transition multigraph (Fig.~\ref{fig:multigraph}); it is succinctly representated by the transition graph of the first program.
Our first contribution (\S\ref{sec:guided_multigraph}) is to recast the ``guided analysis'' from \cite{DBLP:conf/sas/GopanR07} on such a succinct representation of the paths in lieu of the individual transitions.
A similar explosion occurs in disjunctive invariant generation, following \cite{DBLP:conf/pldi/GulwaniZ10}; our second contribution (\S\ref{sec:disjunctive}) applies our implicit representation to their method.

@@

\begin{figure}
  \begin{minipage}{.3\linewidth}
    \lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}
  \end{minipage}
\hfill
\begin{minipage}{.5\linewidth}
\includegraphics[scale=0.7]{gopan_reps}
\vspace{-1em}
\begin{equation}
y \leq x \;\land\; y \leq 102-x \;\land\; y \geq 0.\label{eqn:triangle}
\end{equation}
\end{minipage}
\caption{Example program and its invariant: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
\label{fig:gopan_reps_invariant} 
\end{figure}

@@

Gopan and Reps' idea is to analyze the first phase of the loop with a widening and narrowing sequence, and thus obtain $0 \leq x \leq 50 \land y = x$, and then analyze the second phase, finally obtaining invariant~\ref{eqn:triangle}; each phase is identified by the tests taken or not taken.

@@

Analyzing this loop using widening and narrowing on convex polyhedra or octagons yields the loop invariant $0 \leq x \leq 51 \land y = x$. Now, the transition at line~5 becomes feasible; and we analyze the full loop, starting iterations from $0 \leq x \leq 51 \land y = x$, and obtain invariant~(\ref{eqn:triangle}).

@@

The order of steps is important: narrowing has to be performed before
adding new paths, or some spurious new paths would be added to $P$,
and starting with the addition of new paths avoids wasting time
doing the ascending iterations on an empty graph.

@@

Our improvement consists in a modification of the SMT formula we solve in
\ref{sec:guided_multigraph}.
We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
that we can easily find in the model which abstract value of the
disjunction of the source point has
to be chosen to make the invariant of the destination grow.
The resulting formula that is given to the SMT solver is defined
by $g(p_i)$.
When the formula is satisfiable, we know that the index $j$ of the starting
disjunct that has to be chosen is the one for which the associate Boolean value
$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
$\sigma_i(j,k)$, thus know the index of the disjunct to join with.
%
\begin{equation*}
g(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg b_j^s 
 \wedge 
\displaystyle\bigvee_{1 \leq k \leq m_i}\!\! (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
\neg d_l)
\wedge
\displaystyle\bigvee_{j \in Succ(i)}\!\!
(b_j^d \wedge \bigwedge_{1 \leq k \leq m_i}\!\! (\neg X_{j,k}))
\end{equation*}
%
In our algorithm, the initialization of the abstract values slightly differs from
algorithm \ref{algo:combined} line~\ref{alg=X-init}, since we now have to
initialize each disjunct. Instead of Line~\ref{alg=X-init}, we initialize $X_{i,k}$ with $\perp$ for all
$k \in \{1,..,m_i\} \setminus \{\delta_i\}$, and $X_{i,\delta_i}$ with
$\gets I_{p_i}$.

@@

The $\sigma_i$ functions take as parameters the index of the starting abstract
value, and the path we focus on. 
Since we dynamically construct these functions
during the analysis, we store their already computed image into a
compact representation, such as Algebraic Decision Diagrams.
$\sigma_i(j,k)$ is then constructed on the fly only when needed, and computed
only once.
When the value of $\sigma_i(j,k)$ is required but undefined, we first compute
the image of the
abstract value $X_{i,j}$ by the path indexed by $k$, and try to find an existing
disjunct of index $j'$ so that the least upper bound of the two abstract values is exactly their union (this can be tested using SMT-solving).
If such an index exists, then we set $\sigma_i(j,k) = j'$.
Otherwise:
\begin{compactitem}
	\item if $m_i < M$, we increase $m_i$ by $1$ and define $\sigma_i(j,k) = m_i$
	\item if $m_i = M$, we define $\sigma_i(j,k) = M$ 
		%\JH{A la place, Gulwani cherche l'invariant qui a le plus de contraintes
		%d'egalite en commun pour definir son $sigma$. Nous, on le met dans M, pour simplifier}
\end{compactitem}
The main difference with the original
algorithm~\cite{DBLP:conf/pldi/GulwaniZ10} is that we construct
$\sigma_i(j,k)$ using SMT queries instead of enumerating a possibly
exponential number of paths to find a solution.

@@

We conducted extensive experiments on real-life programs in order to compare the
different techniques, mostly on open-source projects (Fig.~\ref{fig:projects}) written in C, C++ and Fortran.

\begin{table}[!h]
      \centering
      \begin{tabular}{|l|r|r|r|r|r|r|r|} \hline
        \multicolumn{1}{|c|}{} &
        \multicolumn{2}{c|}{Size} &
        \multicolumn{5}{c|}{Execution time (seconds)} \\ \hline
        \multicolumn{1}{|c|}{Name} & 
        \multicolumn{1}{c|}{kLOC}  & 
        \multicolumn{1}{c|}{$|P_R|$}  &
        \multicolumn{1}{c|}{\textbf{S}} &
        \multicolumn{1}{c|}{\textbf{G}} &
        \multicolumn{1}{c|}{\textbf{PF}} &
        \multicolumn{1}{c|}{\textbf{G+PF}} &
	\multicolumn{1}{c|}{\textbf{DIS}} \\ 
        \hline
        % M-x align RET powers ;-).
a2ps-4.14      & 55  & 2012  & 64  & 17  & 66   & 175  & 512   \\
gawk-4.0.0     & 59  & 902   & 59  & 8   & 18   & 43   & 129   \\
gnuchess-6.0.0 & 38  & 1222  & 112 & 21  & 122  & 374  & 866   \\
gnugo-3.8      & 83  & 2801  & 155 & 31  & 128  & 382  & 1403  \\
grep-2.9       & 35  & 820   & 50  & 12  & 21   & 58   & 307   \\
gzip-1.4       & 27  & 494   & 63  & 12  & 36   & 97   & 617   \\
lapack-3.3.1   & 954 & 16422 & 721 & 246 & 2690 & 6052 & 22222 \\
make-3.82      & 34  & 993   & 22  & 7   & 32   & 83   & 255   \\
sed-4.2        & 23  & 292   & 35  & 7   & 23   & 69   & 521   \\
tar-1.26       & 73  & 1712  & 270 & 39  & 110  & 304  & 1164  \\
% TODO: mettre à jour les chiffres ?
	\hline
\end{tabular}
\caption{List of analyzed open-source projects, with their respective number of
lines of code, number of control points in $P_R$, and execution time
on various techniques\MM{TODO}}
\label{fig:projects}
\end{table}

@@

@@

\begin{figure}[h]
  \begin{center}
    \input{gnuplot/techniques.tex}
  \end{center} 
  \vspace{-20pt}
  \caption{Comparison of the abstract values obtained by classical abstract
  interpretation (S), \emph{Guided Static
  Analysis} (G), \emph{Path-focused} technique (PF), our combined technique
  (G+PF), and its version with disjunctive invariants (DIS).
  The $\subsetneq$ bars gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side technique,
$\supsetneq$ the percentage of invariants stronger with the right-side technique,
and ``uncomparable'' gives the percentage of invariants that are uncomparable, i.e
neither greater nor smaller;
the code points where both invariants are equal make up the remaining percentage. See Tab.~\ref{tab:techniques} for details.}
  \label{fig:techniques}
\end {figure}

>>>>

\usepackage{amsmath}
\usepackage{mathcomp,amsfonts,amssymb,mathabx}
\usepackage[mathscr]{euscript}
\usepackage{algorithmic,algorithm}
\usepackage{paralist}
\usepackage{graphicx}
%\usepackage{color}
\usepackage{placeins}

@@

\avirer{
Most SMT solvers, including Z3,
%\footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
Yices,
%\footnote{\url{http://yices.csl.sri.com/}}
and all those supporting the full SMTLIB2 standard \cite{BarST-SMTLIB},
offer an \emph{incremental} interface: the client program specifies the formula as an initially empty conjunction, to which additional constraints are added, and calls a ``check'' function answering whether it is satisfiable; 
it may then backtrack some of the constraints and add other ones without restarting from scratch.
}

@@

Classical interval analysis on this program yields $\lstinline|x_old| \in [-100000,100000]$.
We have transformed the program, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first \avirer{(as in Fig.~\ref{fig:multigraph})}, even though not all feasible paths are needed to compute the invariant.

Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit while simulating its analysis. This means we work on an implicitly represented transition multigraph \avirer{(Fig.~\ref{fig:multigraph})}; it is succinctly representated by the transition graph of the first program.
Our first contribution (\S\ref{sec:guided_multigraph}) is to recast the ``guided analysis'' from \cite{DBLP:conf/sas/GopanR07} on such a succinct representation of the paths in lieu of the individual transitions.
A similar explosion occurs in disjunctive invariant generation, following \cite{DBLP:conf/pldi/GulwaniZ10}; our second contribution (\S\ref{sec:disjunctive}) applies our implicit representation to their method.

@@

\begin{figure}
  \vspace{-1em}
  \begin{minipage}{.3\linewidth}
    \lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}
  \end{minipage}
\hfill
\begin{minipage}{.5\linewidth}
\includegraphics[scale=0.7]{gopan_reps}
\vspace{-1em}
\begin{equation}
y \leq x \;\land\; y \leq 102-x \;\land\; y \geq 0.\label{eqn:triangle}
\end{equation}
\end{minipage}
\caption{Example program and its invariant: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
\label{fig:gopan_reps_invariant} 
\end{figure}

@@

Gopan and Reps' idea is to analyze the first phase of the loop with a widening and narrowing sequence, and thus obtain $0 \leq x \leq 50 \land y = x$, and then analyze the second phase, finally obtaining invariant~(\ref{eqn:triangle}); each phase is identified by the tests taken or not taken.

@@

Analyzing this loop using widening and narrowing on convex polyhedra
or octagons yields the loop invariant $0 \leq x \leq 51 \land y = x$.
Now, the transition at line~5 becomes feasible; and we analyze the
full loop, starting iterations from $0 \leq x \leq 51 \land y = x$,
and obtain invariant~(\ref{eqn:triangle}) in Fig~\ref{fig:gopan_reps_invariant}.

@@

The order of steps is important: narrowing has to be performed before
adding new paths, or spurious new paths would be added to $P$.
Starting with the addition of new paths avoids% wasting time
doing the ascending iterations on an empty graph.

@@

Our improvement consists in a modification of the SMT formula we solve in
\ref{sec:guided_multigraph}.
We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
that we can easily find in the model which abstract value of the
disjunction of the source point has
to be chosen to make the invariant of the destination grow.
The resulting formula that is given to the SMT solver is defined
by $g(p_i)$.
When the formula is satisfiable, we know that the index $j$ of the starting
disjunct that has to be chosen is the one for which the associate Boolean value
$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
$\sigma_i(j,k)$, thus know the index of the disjunct to join with.
%
\vspace{-.5em}
\begin{equation*}
g(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg b_j^s 
 \wedge 
\displaystyle\bigvee_{1 \leq k \leq m_i}\!\! (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
\neg d_l)
\wedge
\displaystyle\bigvee_{j \in Succ(i)}\!\!
(b_j^d \wedge \bigwedge_{1 \leq k \leq m_i}\!\! (\neg X_{j,k}))
\vspace{-.5em}
\end{equation*}

In our algorithm, the initialization of the abstract values slightly differs from
algorithm \ref{algo:combined} line~\ref{alg=X-init}, since we now have to
initialize each disjunct. Instead of Line~\ref{alg=X-init}, we initialize $X_{i,k}$ with $\perp$ for all
$k \in \{1,..,m_i\} \setminus \{\delta_i\}$, and $X_{i,\delta_i}$ with
$\gets I_{p_i}$.

@@

The $\sigma_i$ functions take as parameters the index of the starting abstract
value, and the path we focus on. 
Since we dynamically construct these functions
during the analysis, we store their already computed image into a
compact representation, such as Algebraic Decision Diagrams.
$\sigma_i(j,k)$ is then constructed on the fly only when needed, and computed
only once.
When the value of $\sigma_i(j,k)$ is required but undefined, we first compute
the image of the
abstract value $X_{i,j}$ by the path indexed by $k$, and try to find an existing
disjunct of index $j'$ so that the least upper bound of the two abstract values is exactly their union (using SMT-solving).
If such an index exists, then we set $\sigma_i(j,k) = j'$.
Otherwise:
\begin{compactitem}
	\item if $m_i < M$, we increase $m_i$ by $1$ and define $\sigma_i(j,k) = m_i$
	\item if $m_i = M$, we define $\sigma_i(j,k) = M$ 
		%\JH{A la place, Gulwani cherche l'invariant qui a le plus de contraintes
		%d'egalite en commun pour definir son $sigma$. Nous, on le met dans M, pour simplifier}
\end{compactitem}
The main difference with the original
algorithm~\cite{DBLP:conf/pldi/GulwaniZ10} is that we construct
$\sigma_i(j,k)$ using SMT queries instead of enumerating a possibly
exponential number of paths to find a solution.

@@

We conducted extensive experiments on real-life programs in order to compare the
different techniques, mostly on open-source projects (Fig.~\ref{fig:projects}) written in C, C++ and Fortran.
\MM{TODO: résumer les conclusions sur les résultats}

\begin{figure}[h!]
      \centering
      \begin{tabular}{|l|r|r|r|r|r|r|r|} \hline
        \multicolumn{1}{|c|}{} &
        \multicolumn{2}{c|}{Size} &
        \multicolumn{5}{c|}{Execution time (seconds)} \\ \hline
        \multicolumn{1}{|c|}{Name} & 
        \multicolumn{1}{c|}{kLOC}  & 
        \multicolumn{1}{c|}{$|P_R|$}  &
        \multicolumn{1}{c|}{\textbf{S}} &
        \multicolumn{1}{c|}{\textbf{G}} &
        \multicolumn{1}{c|}{\textbf{PF}} &
        \multicolumn{1}{c|}{\textbf{G+PF}} &
	\multicolumn{1}{c|}{\textbf{DIS}} \\ 
        \hline
        % M-x align RET powers ;-).
a2ps-4.14      & 55  & 2012  & 64  & 17  & 66   & 175  & 512   \\
gawk-4.0.0     & 59  & 902   & 59  & 8   & 18   & 43   & 129   \\
gnuchess-6.0.0 & 38  & 1222  & 112 & 21  & 122  & 374  & 866   \\
gnugo-3.8      & 83  & 2801  & 155 & 31  & 128  & 382  & 1403  \\
grep-2.9       & 35  & 820   & 50  & 12  & 21   & 58   & 307   \\
gzip-1.4       & 27  & 494   & 63  & 12  & 36   & 97   & 617   \\
lapack-3.3.1   & 954 & 16422 & 721 & 246 & 2690 & 6052 & 22222 \\
make-3.82      & 34  & 993   & 22  & 7   & 32   & 83   & 255   \\
sed-4.2        & 23  & 292   & 35  & 7   & 23   & 69   & 521   \\
tar-1.26       & 73  & 1712  & 270 & 39  & 110  & 304  & 1164  \\
% TODO: mettre à jour les chiffres ?
	\hline
      \end{tabular}
      \label{fig:projects}

    \input{gnuplot/techniques.tex}
  \vspace{-20pt}
  \caption{Comparison of the abstract values obtained on several open-source projects. The table shows
  their respective number of lines of code, number of control points in $P_R$, and execution time
  on various techniques.
  Techniques are classical abstract
  interpretation (S), \emph{Guided Static
  Analysis} (G), \emph{Path-focused} technique (PF), our combined technique
  (G+PF), and its version with disjunctive invariants (DIS).
  The $\subsetneq$ bars gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side technique,
$\supsetneq$ the percentage of invariants stronger with the right-side technique,
and ``uncomparable'' gives the percentage of invariants that are uncomparable, i.e
neither greater nor smaller;
the code points where both invariants are equal make up the remaining percentage. See Tab.~\ref{tab:techniques} for details.}
  \label{fig:techniques}
\end {figure}

@@

\FloatBarrier

@@

\MM{TODO: graphique supprimé ici}
